### 2019-06-09 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Done, complete proof-reading of Chapter:4

As explained before, the future Internet of Things (IoT) networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use Multi-Armed Bandit (MAB) learning algorithms to improve spectral resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks.
We mean by \emph{decentralized} that learning is performed on the device side, and is spread on the totality of the devices in a network.
We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms
firstly extends devices battery autonomy (by reducing the packet-loss-ratio or equivalently, by improving the successful-communication rate),
and secondly does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied settings,
Stochastic MAB learning
% provides an up to $16\%$ gain in terms of successful transmission probabilities, and
has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network is less and less stationary.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic in decentralized IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

\textbf{Outline.}
%
The rest of this section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. The reference policies are described in \ref{sub:41:threeReferencePolicies}, and our proposed sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
Numerical results are presented in \ref{sub:41:numericalResults}.

@@

\textbf{One gateway and many devices.}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that devices do not use sensing, and we do not consider retransmissions of packets (we remove this hypothesis in Section~\ref{sec:4:retransmissions}).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption[In our system model, some dynamic devices (in the IoT network in blue) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in orange).]{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange}).}
    \label{fig:41:system_model1}
\end{figure}

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

\paragraph{Slotted protocol.}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement \Ack{} is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

\begin{figure}[!t]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption[The considered time-frequency slotted protocol. Each frame is composed by a fixed duration uplink slot in which the end-devices transmit their (uplink) packets. If a packet is well received, the gateway replies by transmitting an \Ack, after the ack delay.]{The considered time-frequency slotted protocol. Each frame is composed by a fixed duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\paragraph{Static vs dynamic devices.}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\paragraph{Random emission patterns.}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\paragraph{Some assumptions on the network occupation.}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

\paragraph{Link with the MAB model.}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

\textbf{$1^{\text{st}}$ - Naive policy: Random Channel Selection} ``\textcolor{darkgreen}{Random}''.
%
We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{k=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;k)}_{\text{No one else sent in channel}\; k} \; \underbrace{\Pr(\text{sent in channel}\,k)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

\textbf{$2^{\text{nd}}$ - (Unachievable) Optimal oracle policy} ``\textcolor{orange}{Optimal}''.
%
We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of the repartition of static devices $(S_k)_k$, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle chooses a fixed configuration of dynamic devices, with $D_k$ devices affected to channel $k$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \Pr(\text{success}|\text{sent in channel}\;k) \; \Pr(\text{sent in channel}\;k) \nonumber \\
    & = \sum_{k=1}^{K} \underbrace{(1 - p)^{D_k - 1}}_{\;\;D_k - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_k}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_k / D }_{\;\;\text{Sent in channel}\; k\;\;}.
\end{align}

@@

\begin{proposition}\label{prop:41:Lagrangian}
\begin{leftbar}[propositionbar]  % XXX leftbar propositionbar, comment if needed
    The \emph{Lagrange multipliers} method \cite{BoydVanderberghe04} can be used to solve the constraint real-valued maximization problem introduced in equation \eqref{eq:41:prob}.
    %
    It gives a closed form expression for the (unique) optimal solution $D_k^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.
    %
    \begin{equation}\label{eq:41:Dilambda}
        D_k^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda e}{(1-p)^{S_k-1}} \right)-1 \right]\right)^{+} .
    \end{equation}
\end{leftbar}  % XXX leftbar propositionbar, comment if needed
\end{proposition}

\begin{smallproof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_k\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall k\in\llbracket 1;K \rrbracket$. For such values for $D_k$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}$ is concave as the sum of concave functions
    \footnote{~It is worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    % \item
    % \hfill{}$\square$
    \end{itemize}
    More details are given in Section~\ref{sec:4:proofLagrangian} in the Appendix of this Chapter.
\end{smallproof}

>>>>

As explained before in Chapter~\ref{chapter:1}, the future IoT networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use MAB learning algorithms to improve spectral resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks.
We mean by \emph{decentralized} that learning is performed on the device side, and is spread on the totality of the devices in a network.
We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms
firstly extends devices battery autonomy (by reducing the packet-loss-ratio or equivalently, by improving the successful-communication rate),
and secondly does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied settings,
Stochastic MAB learning
% provides an up to $16\%$ gain in terms of successful transmission probabilities, and
has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic in decentralized IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

% \textbf{Outline.}
% %
% The rest of this section is organized as follows. The system model is introduced in Section~\ref{sub:41:systemModel}. The reference policies are described in \ref{sub:41:threeReferencePolicies}, and our proposed sequential learning policies are introduced in \ref{sub:41:sequentialPolicies}.
% Numerical results are presented in \ref{sub:41:numericalResults}.

@@

\textbf{One gateway and many devices.}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that \emph{devices do not use sensing}, and \emph{we do not consider retransmissions of packets} (we relax this hypothesis in Section~\ref{sec:4:retransmissions}).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.70\linewidth]{system_model1.eps}
    \caption[In our system model, some dynamic devices (in the IoT network in blue) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in orange).]{In our system model, some dynamic devices (in the \textcolor{blue}{IoT network in blue}) transmit packets to a gateway and suffer from the interference generated by neighboring networks (in \textcolor{orange}{orange} left/right).}
    \label{fig:41:system_model1}
\end{figure}

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (two are shown in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}, in left or right).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

\paragraph{Slotted protocol.}
%
For simplicity, and for consistency with the rest of this thesis, we only consider a discrete protocol (\ie, slotted).
As illustrated in Figure \ref{fig:41:protocol}, we suppose a slotted protocol, in both time and frequency.
All devices share a synchronized time, and know in advance $K$, the finite number of available RF channels.
In each time slot, the devices try to send packets to the unique gateway, which listens continuously to all channels, following an ALOHA-based communication, with no sensing.
Each time slot is divided in two parts: first for uplink communications in which data packets are sent by end-devices to the gateway.
In one channel, if only one packet is sent in this part of the slot, the gateway can decode it and sends an acknowledgement to the device in the second part (on the same channel).
If two or more devices send an uplink packet in the same slot, the uplink packets collide, and the acknowledgement (\Ack) is not transmitted. In this case, we say that there is a \emph{collision} in this time slot.
In other words, if the gateway cannot successfully decode the incoming (uplink) messages because they were corrupted by collisions, it does not send any \Ack{} back.
This way, no collision can occur on the downlink messages, easing the analysis of collisions.

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.40]{protocol.eps}
    \caption[The considered time-frequency slotted protocol. Each frame is composed by a fixed duration uplink slot in which the end-devices transmit their (uplink) packets. If a packet is well received, the gateway replies by transmitting an \Ack, after the ack delay.]{The considered time-frequency slotted protocol. Each frame is composed by a fixed duration \textcolor{blue}{uplink slot} in which the end-devices transmit their \textcolor{blue}{(uplink) packets}. If a packet is well received, the gateway replies by transmitting an \textcolor{darkgreen}{\Ack}, after the ack delay.}
    \label{fig:41:protocol}
\end{figure}

\textbf{Static vs dynamic devices.}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\textbf{Random emission patterns:}
%
We suppose that all devices follow the same emission pattern, being fixed in time, and we choose to model it as a simple Bernoulli process:
all devices have the same probability to send a packet in any (discrete) temporal slot, and we denote $p \in (0, 1)$ this probability\footnote{~In the experiments below, $p$ is about $10^{-3}$, because in a crowded network $p$ should be smaller than $K / (S + D)$ for all devices to communicate successfully (in average).}.
The parameter $p$ essentially controls the frequency of communication for each device, once the time scale is fixed, and $1/p$ is proportional to the \emph{duty cycle}.
% (\ie, real time during two messages)
For instance, for IoT devices sending messages with a duration of $1$ second and possibly at every second, then on a daily basis we have $p = 1 / (12 \times 60 \times 60) \simeq 1.5 \times 10^{-5}$.

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $D \leq K$ dynamic devices.} than the number of channels}.

\textbf{Link with the MAB model:}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

\paragraph{$1^{\text{st}}$ - Naive policy: Random Channel Selection} (``\textcolor{darkgreen}{Random}'')
%
We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows
%
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{k=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;k)}_{\text{No one else sent in channel}\; k} \; \underbrace{\Pr(\text{sent in channel}\,k)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
%
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

\paragraph{$2^{\text{nd}}$ - (Unachievable) Optimal oracle policy} (``\textcolor{orange}{Optimal}'')
%
We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of the repartition of static devices $(S_k)_k$, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle chooses a fixed configuration of dynamic devices, with $D_k$ devices affected to channel $k$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
%
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \Pr(\text{success}|\text{sent in channel}\;k) \; \Pr(\text{sent in channel}\;k) \nonumber \\
    & = \sum_{k=1}^{K} \underbrace{(1 - p)^{D_k - 1}}_{\;\;D_k - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_k}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_k / D }_{\;\;\text{Sent in channel}\; k\;\;}.
\end{align}

@@

\begin{proposition}\label{prop:41:Lagrangian}
\begin{leftbar}[propositionbar]  % XXX leftbar propositionbar, comment if needed
    The \emph{Lagrange multipliers} method \cite{BoydVanderberghe04} can be used to solve the constraint real-valued maximization problem introduced in equation \eqref{eq:41:prob}.
    %
    It gives a closed form expression for the (unique) optimal solution $D_k^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.
    %
    \begin{equation}\label{eq:41:Dilambda}
        D_k^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda \e}{(1-p)^{S_k-1}} \right)-1 \right]\right)^{+} .
    \end{equation}
\end{leftbar}  % XXX leftbar propositionbar, comment if needed
\end{proposition}
%
\begin{smallproof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_k\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall k\in\llbracket 1;K \rrbracket$. For such values for $D_k$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}$ is concave as the sum of concave functions
    \footnote{~It is worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    % \item
    % \hfill{}$\square$
    \end{itemize}
    More details are given in Section~\ref{sec:4:proofLagrangian} in the Appendix of this Chapter.
\end{smallproof}
